Does at-issueness predict projection? Further investigations of the Gradient Projection Principle

### 2022-02-28 writing/investigating-GPP.tex
updated literature review

\section{Defining and diagnosing at-issueness}\label{s2}

@@

\item Todor Koev

\item Natasha Korotkova

\item Murray, Anderbois et al

\item \citealt{snider2017,snider-paper}

@@

@@

@@

Exp. & Embedding & Projection measure & At-issueness measure \\ 
\hline
1q & question & certain that $c$ & asking whether $c$ \\
1n & negation & certain that $c$ & sure that $c$ \\
1m & modal & certain that $c$ & sure that $c$ \\
1c & conditional & certain that $c$ & sure that $c$ \\
\hline
2q & question & certain that $c$ & {\em yes}, $c$ \\ 
2n & negation & certain that $c$& {\em yes, that's true}, $c$ \\ 
2m & modal & certain that $c$& {\em yes, that's true}, $c$ \\ 
2c & conditional & certain that $c$& {\em yes, that's true}, $c$ \\ 
\hline
3q & question & certain that $c$& {\em yes}, but $\neg c'$ \\ 
3n & negation & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
3m & modal & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
3c & conditional & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
\hline
\end{tabular}
\caption{Overview of the entailment-canceling embeddings, projection measures, and at-issueness measures in Exps.~1-3, where $c$ is the content under investigation, and $c'$ is a salient, different content.}\label{t-overview}
\end{table}

@@

\begin{sidewaystable}[h!]
\centering
\begin{tabular}{l | r r r | r r r r | r r r | r }
&  \multicolumn{3}{c|}{Recruited participants} & \multicolumn{4}{c|}{Exclusion criteria} & \multicolumn{3}{c|}{Remaining participants} &  \\ 
Exp. & total & ages (mean) & f/m/o/u & multiple & language & controls & variance & total & ages (mean) & f/m/o/u & payment   \\ 
\hline
1q & 300  & 19-74 (38.2)  & --  & 5  & 7  & 35  & 0  & 242   & 21-74 (39.2) &  -- & \\
1n & 300  & 18-74 (33.2)  &  150/145/5/0 & 0  & 8 & 17 & 1 & 274  & 18-74 (33.3) & 141/128/5/0 & \$1.70\\
1m & 300  & 18-74 (32.7) & 150/141/7/2  &  0 & 0  & 19 & 0   & 281   & 18-74 (32.7) & 144/129/7/1 & \$1.70 \\
1c &  300 & 18-58 (25.9)  & 249/45/6/0 & 0 & 6  & 26 & 2  & 266  &  18-58 (24.8) & 235/25/6/0  & \$1.70 \\
2q & 250  &  18-58 (25.5) &  201/43/6/0  & 0 & 4  & 24  & 1 & 220  & 18-58 (24.8)  & 187/28/5/0  & \$2.15 \\
2n & 250  &  18-69 (33.2)  &  127/114/6/1 & 1  & 4  & 29 & 0& 215  &  18-69 (33.1) & 113/95/6/1    & \$1.70\\
2m & 251  & 18-74 (31.7)  & 132/113/6/0  & 0  & 4  & 27 &  0 & 220  & 18-70 (31.9) & 116/98/6/0 & \$1.70\\
2c &  250 &  18-56 (24.5)  & 212/30/8/0  & 0  & 0  & 26 & 0  & 224  & 18-56 (24.4) & 195/24/5/0 & \$2.15\\
3q & 250  &  18-66 (32.4) &  140/102/7/1 &  0 & 4  & 20  & 0  &  225 & 18-66 (32.6) & 125/93//7/0  & \$1.70 \\
3n & 250  &  18-70 (24.6) & 114/31/5/0  & 0  & 5  &  13 & 4  & 228  & 18-70 (24.3) &  198/25/5/0 &\$1.70 \\
3m & 250  & 18-63 (25.5)  & 205/40/5/0 &  0 & 3  & 14 &  0 & 233  &  18-63 (24.8) & 197/31/5/0 &\$1.70 \\
3c & 250  & 18-59  (27.5) & 182/64/4/0  & 0  & 3  & 17 & 0 & 230  &  18-59 (26.7)  & 177/49/4/0  & \$2.15\\
\end{tabular}
\caption{Recruited participants, excluded data, and remaining participants in Exps.~1, 2 and 3}\label{f-exclusion}
\end{sidewaystable}

>>>>

\section{Analyzing and diagnosing at-issueness}\label{s2}

\subsection{Formal analyses of at-issueness}

backgrounded content

\begin{exe}
\ex Formal analyses of at-issueness

\begin{xlist}
\ex Dimension-based \\ At-issue content interacts with operators

\ex Question-based \\ At-issue content addresses the question under discussion

\ex Anaphora-based \\ At-issue content introduces a discourse referent for subsequent reference

@@

\item \citealt{potts05}

\item p.23/Figure 2.1: at-issue entailments vs. backgrounded content (conventional presuppositions, CIs, conversationally-triggered presuppositions) vs. conversational implicatures (not backgrounded!!) {\bf 3-way distinction}

\item p.24: But the supplementary relative who lived in a working-class suburb of Boston plays a secondary role relative to the information conveyed by the main clause. The issue is not where the grandmother lived, but rather the fact that the speaker summered with her as a child.

\item p.30: the utterance might be followed in the discourse by ?Hey, she?ll do both!? Or it might be preceded by an agreement that if Mary does one, then she does the other. The maxims of quality and quantity would then conspire to en- sure that (2.35b) disappeared. Thus, the conversational-implicature dimension is a negotiable part of denotations.

\item p.31: a CI is never relativized to the beliefs of an entity other than the speaker. But at-issue content certainly is; in Sue wrongly believes Conner got promoted, the at-issue proposition that Conner got promoted is asserted to hold only in Sue?s belief worlds. this embedded proposition is not speaker-oriented, and hence not classifiable as a CI contri- bution,

\item p.32:  CIs are independent of the at-issue con- tent. In contrast, the fundamental goal of almost all presupposition logics is to create a dependency between the presuppositions and the at-issue entailments

\item p.34: Neither at- issue content nor CI content should be presupposed.

\item p.43: everything about the definition says that CIs contribute new information. In this sense, they are like at-issue content.

\item {\bf can we diagnose not-at-issue content independent of a particular type of not-at-issue content? that is, is there a diagnostic for not-at-issueness that applies to all such content, regardless of its specific type?}

\item p.51: discourse-level phenomena. It seems that none are sensitive to the distinction between these two types of content. (at-issue vs CI): VP ellipsis and do it/that pronominalization

\item p.53: Overall, they are an indication that we distinguish the at-issue and CI dimensions in the composition. From the point of view of a discourse as a whole, they are identical.

\item p.54: A similar, and less fraught, test involves discourse referents. Because noth- ing can scope over supplementary CI content, it is quite good at establishing discourse referents. ... Here again, presupposed material patterns differently. It alone cannot establish a discourse referent, at least not without a significant amount of contextual priming and some creative inferences on the part of the hearer;

\item p.54: Speaker-oriented adverbs, for instance, suggest a new test. Exam- ples (3.16)?(3.17) show that their content can be queried just as at-issue content can.

\item p.54: The upshot of these examples is that CI content is model-theoretically the same as at-issue content. ... Thus, we are led to the con- clusion that the distinction exists only at the level of the meaning language. {\bf potts assumes that ai and nai content are not distinct on a discourse level!}

\item p.70: in the CI logic employed here, translation is a nontrivial step, as it is the locus of the at-issue/CI distinction. If we interpreted natural language expres- sions directly, we would be left with only two options, both undesirable: we could locate the distinction in the natural language syntax or in the models. The syntactic view is initially unpromising given the semantic nature of the defini- tion of CIs, and it receives sustained criticism in chapter 6. The model-theoretic option is to locate the differences in the models. But section 3.3 shows that in- tersentential phenomena treat the two classes of meaning as model-theoretically identical. Moreover, it is hard to see what the distinctions could be.

\item \citealt{potts2007-compass}

\item p.666: ``typically the content that speakers offer as primary and also the content that they are most expecting to have to negotiate with their interlocutors before it is accepted into the common ground''

\item \citealt{koev2013,koev2018,koev2021}

\item \citealt{koev2018}

\item p.1: At-issue content is that part of the sentence meaning that is intuitively felt to express the ?main point? when the sentence is uttered in a given context. Potts (2005; 2007) was among the first to use the term ?at-issue content? in the relevant sense, characterizing it as content that ?carr[ies] the main themes of a discourse? (Potts, 2005, 7), and that speakers are ?most expect- ing to have to negotiate with their interlocutors before it is accepted into the common ground? (Potts, 2007, 666).  The first property, i.e. the ability to contribute to the discourse topic, is often expressed by saying that at-issue content is what is relevant to the question under discussion. The second property, i.e. the ability to represent what is being negotiated among interlocu- tors, is typically captured by saying that at-issue content constitutes a proposal to update the common ground.

\item p.2: There exist several empirical tests for at-issueness, e.g. question/answer pairs and accessibility to direct denials or other discourse continuations. Exist- ing accounts have employed such tests rather indiscriminately or without elaborating on how these tests relate to a given theoretical construal. In order to be able to judge a given theory on its own merits, in each case I will consider only those diagnostics that transparently follow from what counts as at issue in the relevant sense.

\item p.3: q-at-issueness:  A proposition p is Q-atissue relative to a question under discussion Q and a context c iff ? p is relevant to Q in c and
? p is appropriately conventionally marked relative to Q.

\item p.4: Above, Q-at-issueness was diagnosed in question/answer pairs. This is consistent with the question-based definition in 2, provided that an overt question is assumed to manifest the current question under discussion.

\item p.5: several authors identify at-issue content with the proposal introduced by a declarative utterance, i.e. with its asserted content (AnderBois et al., 2015; Farkas \& Bruce, 2010; Koev, 2013; Murray, 2014; Potts, 2005). {\bf potts 2005 does not assume that CIs are not asserted!}

\item p.5: A proposition p is P-at-issue in a context c iff ? p is a proposal in c and ? p has not been accepted or rejected in c.

\item p.5: to test a proposition for proposalhood, we need a lin- guistic tool that tracks what is being negotiated at any given point. One widely used diagnostic to this effect is the assent/dissent test, which probes the amenability of semantic content to direct responses

\item p.6: This test brings up the question of what counts as a direct response. One peculiarity about the responses in 10?11 is that they are anaphoric to parts of the target utterance through propositional anaphors like that or no.7 This is important because nonanaphoric denials behave differently: They can felicitously target nominal appositives (see also Hunter & Asher, 2016).

\item p.6: One possible reaction to the contrast in 10?11 v. 12?13 is to say that the assent/dissent test tracks anaphoric potential rather than P-at-issueness. This is reasonable, although it has also been claimed that P-at-issue content obligatorily establishes a propositional discourse referent (Murray, 2014).

\item p.6: I will thus follow the bulk of the literature in assuming that such responses target the assertion and will employ them as the main diagnostic for P-at-issueness, leaving the precise link between anaphoricity and at-issueness to further research.

\item p.7: A third notion of at-issueness is situated within theories which view discourse as structured by coherence relations, also called ?rhetorical? or ?discourse? relations (Asher \& Lascarides, 2003; Hobbs, 1979, 1985; Kehler, 2002; Mann & Thompson, 1988).

\item p.8:  A proposition expressed by a node ?? of a discourse tree D is C-atissue iff ?? is on the right frontier of D.

\item p.8: A proposition is C-atissue iff a freshly uttered segment can attach to it by some appropriate coherence relation.

\item p.9: Coherence-based accounts thus view the assent/dissent test as a special case of discourse attachment, i.e. one in which the response is connected to the target by an Agreement/Denial relation.

\item p.9: Q-at-issueness is primarily a backward-looking notion or relates to existing discourse. By contrast, P-at-issueness and C-at-issueness are forward-looking or relate to what is to come.

\item p.10: I first compare Q-at-issueness and P-at-issueness, arguing that neither property entails the other.

\item p.10: Koev argues that 26 shows that the appositive is not P-at-issue but that 27 shows that it is Q-at-issue: ``appositive relative clauses are often taken into consideration when answering the question under discussion. Here, the crucial information that Mary experienced firsthand the terrible consequences of cancer comes from the appositive; the main clause alone does not answer the question, at least not without loss of information.''

(26) A: Mary took care of her husband, who had prostate cancer, for almost a year. B: ? No, he didn't.

(27) Q: Why is Mary fundraising for the upcoming Walk For Cancer?
A: She took care of her husband, who had prostate cancer, for almost a year.

\item p.10: Relevance implicatures. They are not P-at-issue (as diagnosed by dissent) but Q-at-issue (as diagnosed by Q/A pairs).

\item p.10f: slifting constructions. Can be P-at-issue (dissented with) but don't answer Q (diagnosed by Q/A pairs)

\item p.11:  Simons et al. (2010) and Beaver et al. (2017) claim that projectivity is perfectly correlated with (the lack of) Q-at-issueness {\bf NOT TRUE}.

\item p.12: The Projection Principle then appears to be problematic in both directions: Projecting content can be Q-at issue, and nonprojecting content can be not Q-at issue.  {\bf Need to engage with Koev 2018 as somebody who doesn't believe in the GPP}

\item p.12: The second observation is that the Projection Principle is stated in terms of Q-at-issueness {\bf NO!} and other notions of at-issueness may lead to different predictions. The question of which notion of at-issueness (if any) makes the right predictions about projection is yet to be systematically investigated.

\item endnote 1: I also limit my attention to declarative sentences. The question of how or whether at-issueness interacts with other sentence types will have to await another study.

\item endnote 5: The assent/dissent test is known under different names in the literature and has been used to distinguish asserted content from implications triggered by constructions as different as presupposition markers (Chierchia & McConnell-Ginet, 2000; Karttunen & Peters, 1979; Shanon, 1976; Strawson, 1950), epistemic modals (Lyons, 1977; Papafragou, 2006; von Fintel & Gillies, 2007), evidentials (Faller, 2002; Matthewson et al., 2007; Murray, 2014), and appositives (AnderBois et al., 2015; Koev, 2013; Syrett & Koev, 2015; Tonhauser, 2012).

\item endnote 16: Appositive content typically projects, so it could be argued that it is not ?intended? as being at issue (see Simons et al., 2010).
However, given its significance in the above context and the fact that conventional marking of at-issueness is often overruled, the information that the director fired Jon would very likely be taken into consideration when answering the question under discussion.

\item \citealt{amaral-etal2007}

\item \citealt{gutzmann}

\item \citealt{korotkova2020}

\item \citealt{murray2014}

\item \citealt{anderbois-etal2015}

\item \citealt{snider-thesis, snider2017, snider2018}

@@

\item \citealt{brst-salt10}

\item \citealt{brst-ar}

\subsection{Diagnosing at-issueness}

@@

\subsection{Next steps}

@@

{\bf Exp.} & {\bf Embedding} & {\bf Projection measure} & {\bf At-issueness measure} \\ 
\hline
1q & question & certain that $c$ & asking whether $c$ \\
1n & negation & certain that $c$ & sure that $c$ \\
1m & modal & certain that $c$ & sure that $c$ \\
1c & conditional & certain that $c$ & sure that $c$ \\
\hline
2q & question & certain that $c$ & {\em yes}, $c$ \\ 
2n & negation & certain that $c$& {\em yes, that's true}, $c$ \\ 
2m & modal & certain that $c$& {\em yes, that's true}, $c$ \\ 
2c & conditional & certain that $c$& {\em yes, that's true}, $c$ \\ 
\hline
3q & question & certain that $c$& {\em yes}, but $\neg c'$ \\ 
3n & negation & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
3m & modal & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
3c & conditional & certain that $c$& {\em yes, that's true}, but $\neg c'$ \\ 
\hline
\end{tabular}
\caption{Overview of the entailment-canceling embeddings, projection measures, and at-issueness measures in Exps.~1-3, where $c$ is the content under investigation (in the target trials: the content of the clausal complement) and $c'$ is a different, salient content (in the target trials: the main clause content).}\label{t-overview}
\end{table}

@@

\begin{sidewaystable}[h!]
\centering
\begin{tabular}{l | r r r | r r r r | r r r | r }
&  \multicolumn{3}{c|}{Recruited participants} & \multicolumn{4}{c|}{Exclusion criteria} & \multicolumn{3}{c|}{Remaining participants} &  \\ 
Exp. & total & ages (mean) & f/m/o/u & multiple & language & controls & variance & total & ages (mean) & f/m/o/u & payment   \\ 
\hline
1q & 300  & 19-74 (38.2)  & --  & 5  & 7  & 35  & 0  & 242   & 21-74 (39.2) &  -- & \$1.70\\
1n & 300  & 18-74 (33.2)  &  150/145/5/0 & 0  & 8 & 17 & 1 & 274  & 18-74 (33.3) & 141/128/5/0 & \$1.70\\
1m & 300  & 18-74 (32.7) & 150/141/7/2  &  0 & 0  & 19 & 0   & 281   & 18-74 (32.7) & 144/129/7/1 & \$1.70 \\
1c &  300 & 18-58 (25.9)  & 249/45/6/0 & 0 & 6  & 26 & 2  & 266  &  18-58 (24.8) & 235/25/6/0  & \$2.15 \\
2q & 250  &  18-58 (25.5) &  201/43/6/0  & 0 & 4  & 24  & 1 & 220  & 18-58 (24.8)  & 187/28/5/0  & \$1.70 \\
2n & 250  &  18-69 (33.2)  &  127/114/6/1 & 1  & 4  & 29 & 0& 215  &  18-69 (33.1) & 113/95/6/1    & \$1.70\\
2m & 251  & 18-74 (31.7)  & 132/113/6/0  & 0  & 4  & 27 &  0 & 220  & 18-70 (31.9) & 116/98/6/0 & \$1.70\\
2c &  250 &  18-56 (24.5)  & 212/30/8/0  & 0  & 0  & 26 & 0  & 224  & 18-56 (24.4) & 195/24/5/0 & \$2.15\\
3q & 250  &  18-66 (32.4) &  140/102/7/1 &  0 & 4  & 20  & 0  &  225 & 18-66 (32.6) & 125/93//7/0  & \$1.70 \\
3n & 250  &  18-70 (24.6) & 114/31/5/0  & 0  & 5  &  13 & 4  & 228  & 18-70 (24.3) &  198/25/5/0 &\$1.70 \\
3m & 250  & 18-63 (25.5)  & 205/40/5/0 &  0 & 3  & 14 &  0 & 233  &  18-63 (24.8) & 197/31/5/0 &\$1.70 \\
3c & 250  & 18-59  (27.5) & 182/64/4/0  & 0  & 3  & 17 & 0 & 230  &  18-59 (26.7)  & 177/49/4/0  & \$2.15\\
\end{tabular}
\caption{Recruited participants, excluded data, and remaining participants in Exps.~1, 2 and 3}\label{f-exclusion}
\end{sidewaystable}
